---
title: "My notes"
author: "Akanksha"
date: "5/6/2019"
output: html_document
---
```{r}
install.packages("nycflights13")
library(nycflights13)
library(readr)
library(tidyverse)

tidyverse_packages() ## lists all the packages that come with the tidyverse package - those not loaded automatically have to be loaded separately.

dta <- read_csv(here("/03-shell/examples/meals","friday.csv"))
rm(data,dta)
```

```{r Filter Command}
starwars
starwars %>% 
  dplyr::filter(skin_color=="white") %>% 
  filter(eye_color =="yellow")

starwars %>% 
  filter(skin_color=="white", eye_color =="yellow") 

starwars %>% 
  filter(grepl("Vader",name))

starwars %>% 
  select(name, height) %>%
  filter(name %in% c("Skywalker"))

starwars %>% 
  select(name, height) %>%
  filter(name %in% c("Luke Skywalker")) %>% 
  
## Useful in excluding NA

starwars %>% 
  filter(is.na(hair_color))

starwars %>% 
  filter(!is.na(hair_color))

```

```{r Arrange Command}

starwars %>% 
  filter(skin_color=="fair") %>% 
  dplyr::arrange(height)

starwars %>% 
  filter(skin_color=="fair") %>% 
  dplyr::arrange(desc(height))

starwars %>% 
  filter(skin_color=="fair") %>% 
  dplyr::arrange(desc(name))

```

```{r Select Command and Renaming Variables}
starwars %>% 
  select(mass:gender, -skin_color, films)

starwars %>%
  select(alias=name, crib=homeworld, sex=gender)  ## can rename variables this way 

## otherways to rename variables

z <- list(a = 1, b = "c", c = 1:3)
names(z)
# change just the name of the third element.
names(z)[3] <- "c2"
z

z <- 1:3
names(z)
## assign just one name
names(z)[2] <- "b"
z

names(starwars) <- toupper(names(starwars))
names(starwars) <- tolower(names(starwars))

starwars %>% 
  select(mass, contains("es"))

## reordering of variables ------------
starwars %>% 
  select(name, contains("color"),everything()) 

```

```{r Mutate Command}
starwars %>% 
  select(name,hair_color,birth_year) %>% 
  mutate(comment = paste0("The color of the hair of ", name, " is ", hair_color))

starwars %>% 
  select(name, birth_year) %>%
  mutate(dog_years = birth_year * 7) %>%
  mutate(comment = paste0(name, " is ", dog_years, " in dog years."))
## Alternatively
starwars %>% 
  select(name, birth_year) %>%
  mutate(
    dog_years = birth_year * 7, ## Separate with a comma
    comment = paste0(name, " is ", dog_years, " in dog years.")
    )


## Mutate all, Mutate if and Mutate at ----------------
# Taking from the exammples given in R

iris <- as_tibble(iris)

iris %>% 
  mutate_if(is.factor,as.character) %>% 
  mutate_if(is.double,as.integer)

```

```{r Summarise Command}
starwars %>% 
  group_by(species,gender) %>% 
  summarise(mean_height = mean(height, na.rm = TRUE))

## Summarise all, Summarise if and Summarise at ----------------
starwars %>% 
  select(height, mass, birth_year ) %>% 
  summarise_all(list(min, max,mean),na.rm=T)

## better to use the funs command instead of list
summ <- starwars %>% 
  select(height, mass, birth_year ) %>% 
  summarise_all(funs(avrg = mean, min=min, maxm = max),na.rm=T)

starwars %>%
  summarise_at(c("height", "mass"), mean, na.rm = TRUE)

starwars %>%
  summarise_at(vars(height:hair_color), mean, na.rm = TRUE)

starwars %>% 
  group_by(species, hair_color, gender) %>% 
  summarise_if(is.numeric, funs(avg=mean, min=min),na.rm=T) %>% 
  head(10)
```

<<<<<<< HEAD
```{r Other Dplyr Goodies - Pull, Slice, (Un)Group, Count and Distinct}

## group and ungroup ----------------
by_cyl <- mtcars %>% group_by(cyl)

by_cyl %>% 
  summarise(mean_mpg = mean(mpg, na.rm = T))

mtcars %>% 
  summarise(mean_mpg = mean(mpg, na.rm = T))

by_cyl %>% 
  ungroup() %>% 
  summarise(mean_mpg = mean(mpg, na.rm = T))

## slice ----------------

starwars %>% 
  slice(1:10) %>% 
  summarise_if(is.numeric, funs(avg=mean, min=min),na.rm=T)
  
## pull ----------------
height_fem <- starwars %>% 
  filter(gender=="female") %>% 
  pull(height)

# count and distinct ----------------
starwars %>% 
   count(gender)

starwars %>% 
  filter(name=="Luke Skywalker")

starwars %>% 
   distinct(gender)

starwars %>% 
  group_by(species) %>% 
  mutate(num = n()) %>% 
  mutate(height_avr = mean(height,na.rm = T)) %>% 
  select(name,num,species,height_avr, contains("color"),everything())
```
```{r Window Functions - Ranking and leads and lags}

## Going through the Vignette of Window Functions -------------------------------------------------------
library(Lahman)
batting <- Lahman::Batting %>% 
  select(playerID, yearID, teamID, G, AB:H) %>% 
  arrange(playerID, yearID, teamID) %>%
  semi_join(Lahman::AwardsPlayers, by= "playerID")

##semi_join() - return all rows from x where there are matching values in y, keeping just columns from x.

players <- batting %>% 
   group_by(playerID)

players %>% 
  mutate(minrank = min_rank(desc(H)))

batting %>% 
  mutate(minrank = min_rank(desc(H)))

batting %>% 
  count(playerID)

# For each player, find the two years with most hits
players %>% 
  filter(min_rank(desc(H)) <= 2 & H > 0)
# Within each player, rank each year by the number of games played
players %>% 
  mutate(rank_g = min_rank(G))

# For each player, compute avg change in games played per year
mutate(players, G_change = (G - lag(G)) / (yearID - lag(yearID)))

players %>% 
  mutate(improve = ifelse(G>lead(G),1,0)) 

##lead() and lag() have an optional argument order_by. If set, instead of using the row order to determine which value comes before another, they will use another variable. This is important if you have not already sorted the data, or you want to sort one way and lag another.

# For each player, find all where they played more games than average
filter(players, G > mean(G))
# For each, player compute a z score based on number of games played
mutate(players, G_z = (G - mean(G)) / sd(G))


players %>% 
  mutate(per=percent_rank(H)) %>% 
  mutate(cper=cume_dist(H)) 

##ntile() divides the data up into n evenly sized buckets. It’s a coarse ranking, and it can be used in with mutate() to divide the data into buckets for further summary. For example, we could use ntile() to divide the players within a team into four ranked groups, and calculate the average number of games within each group.

by_team_player <- group_by(batting, teamID, playerID)
by_team <- summarise(by_team_player, G = sum(G))
by_team_quartile <- group_by(by_team, quartile = ntile(G, 4))
summarise(by_team_quartile, mean(G))

df <- data.frame(year = 2000:2005, value = (0:5) ^ 2)
df
scrambled <- df[sample(nrow(df)), ]
scrambled
```

```{r  Window Functions - Aggregation functions}

## https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html

```

```{r Other Dplyr Goodies - Join Operators}
library(nycflights13)
flights 
planes

left_join(flights, planes) %>%
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, type, model)

flights %>% 
  rename(year_built = year) %>% 
  left_join(planes, by = "tailnum") %>%  ## Be specific about the joining column
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, year_built, type, model) %>%
  head(3) 

## or

left_join(
  flights,
  planes %>% rename(year_built = year), ## Not necessary w/ below line, but helpful
  by = "tailnum" ## Be specific about the joining column
  ) %>%
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, year_built, type, model) %>%
  head(3) 

## or
left_join(
  flights,
  planes, ## Not renaming "year" to "year_built" this time
  by = "tailnum"
  ) %>%
  select(contains("year"), month, day, dep_time, arr_time, carrier, flight, tailnum, type, model) %>%
  head(3)

## for more - https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html
```

```{r tidyr::gather, spread and separate}

stocks <- data.frame( ## Could use "tibble" instead of "data.frame" if you prefer
  time = as.Date('2009-01-01') + 0:1,
  X = rnorm(2, 0, 1),
  Y = rnorm(2, 0, 2),
  Z = rnorm(2, 0, 4)
  )

stocks %>% gather(key=stock, value=price, -time)

# get first observation for each Species in iris data -- base R
mini_iris <- iris[c(1, 51, 101), ]
# gather Sepal.Length, Sepal.Width, Petal.Length, Petal.Width
gather(mini_iris, key = "flower_att", value = "measurement",
       Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)
# same result but less verbose
gather(mini_iris, key = "flower_att", value = "measurement", -Species)

# repeat iris example using dplyr and the pipe operator
mini_iris <-
  iris %>%
  group_by(Species) %>%
  slice(1)
mini_iris %>% gather(key = "flower_att", value = "measurement", -Species)

data <- data.frame(
  village_name = c("a","b","c","d"),
  state_name = rep("mp",times=4),
  pop_1991 =c(100,10,90,89),
  pop_2001 = c(20,90,80,78),
  lit_1991 =c(0.7,0.9,0.45,0.34),
  lit_2001 = c(0.8,0.45,0.12,0.26)
)
data

data %>% 
  gather(key=attribute_yr, value, pop_1991:lit_2001)

data %>% 
  gather(key=attribute_yr, value, pop_1991:lit_2001) %>% 
  separate(attribute_yr, c("col","year"))

data %>% 
  gather(key=attribute_yr, value, pop_1991:lit_2001) %>% 
  separate(attribute_yr, c("col","year")) %>% 
  arrange(village_name)

data %>% 
  gather(key=v, value, pop_1991:lit_2001) %>% 
  separate(v, c("col", "year")) %>% 
  arrange(village_name) %>% 
  spread(col, value)


```

```{r Expand and Complete}
mtcars

tidyr::expand(mtcars, vs, cyl)
tidyr::expand(mtcars, nesting(vs, cyl))

# Implicit missings ---------------------------------------------------------
df <- tibble(
  year   = c(2010, 2010, 2010, 2010, 2012, 2012, 2012),
  qtr    = c(   1,    2,    3,    4,    1,    2,    3),
  return = rnorm(7)
)
df
df %>% tidyr::expand(year, qtr)
df %>% tidyr::expand(year = 2010:2012, qtr)
df %>% tidyr::expand(year = full_seq(year, 1), qtr)
df %>% complete(year = full_seq(year, 1), qtr)

# Nesting -------------------------------------------------------------------
# Each person was given one of two treatments, repeated three times
# But some of the replications haven't happened yet, so we have
# incomplete data:
experiment <- tibble(
  name = rep(c("Alex", "Robert", "Sam"), c(3, 2, 1)),
  trt  = rep(c("a", "b", "a"), c(3, 2, 1)),
  rep = c(1, 2, 3, 1, 2, 1),
  measurment_1 = runif(6),
  measurment_2 = runif(6)
)
experiment

# We can figure out the complete set of data with expand()
# Each person only gets one treatment, so we nest name and trt together:
all <- experiment %>% tidyr::expand(nesting(name, trt), rep)
all

# We can use anti_join to figure out which observations are missing
all %>% anti_join(experiment) ## anti_join(x, y) drops all observations in x that have a match in y.

# And use right_join to add in the appropriate missing values to the
# original data
experiment %>% right_join(all)
# Or use the complete() short-hand
experiment %>% complete(nesting(name, trt), rep)

```

```{r %in% and other built in constants}
1:10 %in% c(1,3,5,9)

sstr <- c("c","aa","B","bba","c",NA,"@","bla","a","Ba","%")
sstr %in% c(letters, LETTERS)
sstr[sstr %in% c(letters, LETTERS)]
```

